### 2014-10-07 Paper_v1.tex
First complete draft of the introduction

\section*{Introduction}
Developing mathematical models of biochemical networks is a significant facet of systems biology.
Modeling approaches differ in their degree of detail, and the choice of approach is often determined by prior knowledge, or model requirements \citep{2012_kholodenko_kolch_SciSig}. 
Ordinary differential equation (ODE) models are a common tool for modeling biochemical systems because of their ability to capture dynamics. 
However, ODEs often come with difficult (and sometimes impossible) parameter identification problems. 
For example, Gadkar et al., showed that even with near-perfect information and high frequency sampling, 
it was often impossible to identify (to within a specified uncertainty) all the parameters in typical signal transduction models \citep{Gadkar:2005ad}. 
ODE models can also require significant mechanistic knowledge of the underlying biology. 
However, Bailey suggested more than a decade ago, that achieving qualitative or even quantitative understanding of biological systems 
should not require complete structural and parametric knowledge \citep{2001_bailey_NatBiotech}. 
Since Bailey's complex biology with no parameters hypothesis, Sethna and coworkers have shown that model behavior and predictive ability is typically 
sensitive to only a few parameter combinations, a characteristic seemingly universal to multi-parameter signaling models referred to as \textit{sloppiness} \citep{Machta:2013aa}.
Thus, accurate model predictions from a sloppy model may be possible, even if the parameters or mechanistic structure is not well understood, 
if the critical directions in parameter space are properly constrained. 
Alternatively, data-driven approaches \citep{Cirit:2012kx} or logical model formulations \citep{Morris:2011ys} 
are emerging paradigms that constrain model structure (and hence the identifiability challenge) by the availability of training and validation data. 
These are interesting techniques because the data naturally informs model structure, which can be expanded as more data becomes available. 
While not mechanistic, these models can simulate surprisingly rich system responses in the absence of mechanism [REFHERE]. 
However, modeling dynamics using logical models is not straightforward. 
Thus, a potential third approach could be a combination of ODEs and logical models, where mechanistic information was used when available, 
and missing or incomplete knowledge encoded using causal local rules.

@@

There are many mathematical models of the human coagulation cascade in the literature.
Previous coagulation models have typically been formulated as systems of nonlinear ordinary differential equations, 
often using mass-action kinetics to describe the rates of biochemical conversions [REFHERE].
Previous ODE coagulation models from our laboratory \citep{Luan:2007aa,2010_luan_varner_MolBioSys}, 
were built upon the pioneering studies of Jones and Mann \citep{Jones:1994aa}, 
and later Hockin \emph{et al.} \citep{Hockin:2002aa}, who developed and subsequently refined a highly mechanistic model of the extrinsic coagulation cascade. 
Other laboratories have also expanded upon Hockin \emph{et al.}, for example Chatterjee \emph{et al.,} augmented the Hockin model to include 
the intrinsic pathway, and the role of thrombin in clot formation \citep{Chatterjee:2010aa}. 
More recently, modeling studies have explored other biochemical or biophysical aspects of coagulation, such as the biochemical networks inside platelets \citep{Stalker:2013aa}, 
multi-scale models of clot formation \citep{Leiderman:2014aa, Bannish:2014ab}, and the transport of material inside growing clots \citep{Voronov:2013aa}.
However, most of these previous studies share the common feature of a mechanism based description of coagulation. 
Such a description is possible because blood, while enormously complex, can be systematically interrogated. 
However, other biochemical networks, such as the intracellular networks which mediate cancer, are more difficult to probe and model.  
Towards this need, we have proposed a hybrid approach which combined ODEs and logical rules that can be used to model 
biochemical processes where mechanistic information is difficult or impossible to obtain. 
We have used this approach to model the coagulation cascade, because of the wealth of training and validation data available for this system.
Our hypothesis is that coagulation represents a best case, if our approach fails in this proof-of-concept study, it will never be acceptable in model difficult cases.
We proposed a reduced order coagulation model that consisted of five differential equations augmented with several logical rules.
We identified the reduced order coagulation model parameters from in vitro extrinsic data sets of thrombin generation initiated by TF/FVIIa in the presence of ATIII, with and without the protein C pathway. We then validated the model using thrombin data sets from multiple laboratories, for both normal and hemophilia coagulation.
The reduced order hybrid approach produced a surprisingly predictive coagulation model, suggesting this framework could potentially be used to model other biochemical networks
important to human health.

>>>>

\section*{Introduction}
Developing mathematical models of biochemical networks is a significant facet of systems biology.
Modeling approaches differ in their degree of detail, and the choice of approach is often determined by prior knowledge, or model requirements \citep{2012_kholodenko_kolch_SciSig}. 
Ordinary differential equation (ODE) models are a common tool for modeling biochemical systems because of their ability to capture dynamics. 
However, ODEs often come with difficult (and sometimes impossible) parameter identification problems. 
For example, Gadkar et al., showed that even with near-perfect information, 
it was often impossible to identify all the parameters in typical signal transduction models \citep{Gadkar:2005ad}. 
However, Bailey suggested more than a decade ago, that achieving qualitative or even quantitative understanding of biological systems 
should not require complete structural and parametric knowledge \citep{2001_bailey_NatBiotech}. 
Since Bailey's complex biology with no parameters hypothesis, Sethna and coworkers have shown that model behavior and predictive ability is typically 
sensitive to only a few parameter combinations, a characteristic seemingly universal to multi-parameter models referred to as \textit{sloppiness} \citep{Machta:2013aa}.
Thus, reasonable predictions from sloppy models may be possible, even if the parameters or structure is uncertain, 
if a few critical directions in parameter space are properly constrained. 
Typical ODE models can also require significant mechanistic information.
Alternatively, data-driven approaches \citep{Cirit:2012kx} or logical model formulations \citep{Morris:2011ys} 
are emerging paradigms that constrain model structure by the availability of training and validation data. 
These are interesting techniques because the data naturally informs model structure, which can be expanded as more data becomes available, and limits the 
identification challenge. Although not mechanistic, these models can simulate surprisingly rich system responses in the absence of mechanism [REFHERE]. 
However, modeling dynamics using logical models is not straightforward. 
Thus, a potential third approach could be a combination of ODEs and logical models, where mechanistic information was used when available, 
and missing or incomplete knowledge encoded using causal local rules.

@@

Coagulation models in the literature have typically been formulated as systems of nonlinear ordinary differential equations, 
often using mass-action kinetics to describe the rates of biochemical conversions [REFHERE].
Previous ODE coagulation models from our laboratory \citep{Luan:2007aa,2010_luan_varner_MolBioSys}, 
were built upon the earlier studies of Jones and Mann \citep{Jones:1994aa}, 
and later Hockin \emph{et al.} \citep{Hockin:2002aa}, who developed and then refined a highly mechanistic model of extrinsic coagulation. 
Other laboratories have also expanded upon Hockin \emph{et al.}, for example including the intrinsic pathway, and thrombin mediated clot formation \citep{Chatterjee:2010aa}. 
Other biochemical or biophysical aspects of coagulation have also been modeled, such as platelet biochemistry \citep{Stalker:2013aa}, 
multi-scale models of clot formation \citep{Leiderman:2014aa, Bannish:2014ab}, and transport inside growing clots \citep{Voronov:2013aa}.
However, most of these previous studies were based upon detailed mechanistic knowledge. 
Such a description is possible because blood, while enormously complex, can be systematically interrogated. 
However, other systems, such as the intracellular networks which mediate cancer, are more difficult to experimentally probe and model.  
Towards this unmet need, we have formulated a hybrid modeling approach which combines ODEs and logical rules to model 
biochemical processes for which mechanistic information is difficult or impossible to obtain. 
We have used this approach to model the coagulation cascade because of the wealth of training and validation data available for this system. 
We proposed a reduced order coagulation model that consisted of five differential equations augmented with several logical rules.
We identified the reduced order coagulation model parameters from in vitro extrinsic coagulation data sets, in the presence of ATIII, with and without the protein C pathway. 
We then validated the model using thrombin data sets from multiple laboratories, for both normal and hemophilic coagulation.
Flux and sensitivity analysis identified which parameter or parameters combinations were critical to model performance in several conditions. 
The reduced order hybrid approach produced a surprisingly predictive coagulation model, suggesting this framework could potentially be used to model other biochemical networks important to human health.
